Hyperplane Elimination for Quickly Enumerating Local Optima

### 2015-09-15 Hyperplane Enumeration/main.tex
Edit pass of the second half

@@

\subsection{$k$-Bound Separable Problem}
A $k$-bound separable problem any problem that is composed of non-overlapping
subfunctions, each using using $k$ or fewer bits. Formally this means
$\forall_i |s_i| \leq k$ and $\forall_{i \neq j} s_i \cap s_j = \emptyset$.
Regardless
of how they are initially ordered, reordering ensures that
all variables which appear in the same $s_i$ are consecutively placed in enumeration.
Consider a numbering of $f_i$ such that if $i < j$ then all of $s_i$'s
variables appear before $s_j$'s in enumeration ordering.

Algorithm~\ref{alg-enumerate} finds the $f_i$ with the highest $i$ which contains a
fitness-improving move and enumerates its variables until it no longer
contains an improving move. In total, enumerating $f_i$ requires less than $2^k$
steps. Each $f_i$ is considered sequentially until none contains an improving move,
meaning the first local optimum is found in \BigO{2^kN/k} time.

Once the first local optimum is found, Algorithm~\ref{alg-enumerate} proceeds
by finding all local optimum in larger and larger hyperplanes. Initially $f_0$
is enumerated to find all local optima in the hyperplane with all $f_i, i>0$
fixed. This process ends when a carry operation modifies a bit of $f_1$. At
this point $f_1$ is enumerated, such that each time $f_1$ contains no improving moves
$f_0$ is enumerated again. This finds all local optima in the hyperplane with $f_i, i>1$
fixed. We represent the time required to find all local optima in the hyperplane
with $f_j, j>i$ fixed as $T(i)$. $T(0)=2^k$ as all values of $f_0$ must be enumerated.
$T(i) = |l_i|*T(i-1)+2^k$ where $|l_i|$ is the number of ways $f_i$ can be set such
that it contains no fitness-improving moves. Each $l_i$ exposes a new hyperplane
that can contain local optima, which causes the recursive call to $T(i-1)$.
Assuming for simplicity that all subfunctions use exactly $k$ bits,
the time required to find all local optima is $T(N/k-1)$.
Assuming that $\forall_i |l_i|=c$, $T(N/k-1)=\sum_0^{N/k-1}2^kc^i<2^kc^{N/k}$.
The number of local optima in the landscape is $c^{N/k}$ meaning that Algorithm~\ref{alg-enumerate}
is within a constant factor ($2^k$) of optimal.

@@

@@

\begin{table}
	\centering
	\caption{Value of $m$ when fitting a model to $y = 2^c2^{mN}$ where $y$ is number of seconds required to
	         find all 1-bit local optima of a problem using $N$ bits. Optimal refers to the same model where
	         $y$ is the number of local optima.}
	\begin{tabular}{|r|c|c|c|c|}
	  \hline
	    & \textbf{Gray-Box} & \textbf{Hyper} & \textbf{Hyper-Reorder} & \textbf{Optimal} \\ \hline
    Concatenated Traps & 0.9809 & 0.5831 & 0.1620 & 0.2000 \\ \hline
    Adjacent NKq & 0.9922 & 0.5972 & 0.3419 & 0.3603 \\ \hline
    Random NKq & 0.9920 & 0.6229 & 0.3989 & 0.3531 \\ \hline
    Ising Spin Glass & 0.9821 & 0.6722 & 0.5976 & 0.6015 \\ \hline
    MAX-kSAT & 0.9949 & 0.8922 & 0.7737 & 0.5393 \\ \hline
  \end{tabular}
  \label{table-scaling}
\end{table}

With the exception of MAX-kSAT, Hyper-Reorder's time required to find all local optima
appears to be growing at the same rate as the total number of local optima. We suspect
some of this deficiency on MAX-kSAT is due to its significantly higher number of non-linear
interactions. Concatenated Traps, Adjacent NKq, and Ising Spin Glass all have exactly 4
non-linear relationships per variable. For small problem sizes
Random NKq and MAX-kSAT both increase with problem
size. Random NKq's average ranges between 5 and 6 on problems we tested.
With $N=15$, MAX-kSAT has on average nearly
12 non-linear relationships per variable. By $N=30$ that grows to just over 17. Having each
variable depend on over half the genome significantly limits the size of hyperplanes which
can be eliminated. While the number of non-linear relationships per variable
is asymptotically constant, this growth on small problems is likely affecting runtime.

\subsection{Finding $r$-Bit Local Optima}
A major advantage of Gray-Box enumeration is that it can find $r$-bit local optima
without exponentially increasing runtime complexity. Figure~\ref{fig-slope-method}
shows the estimated growth rate for each method as $r$ is increased. On Concatenated Traps,
Adjacent NKq, Ising Spin Glass, and to a lesser extent Random NKq, none of our enumeration
methods see an increase in growth complexity with increasing $r$. In all cases using Hyper-Reorder
has a lower growth rate than the other two methods using the same $r$. On every problem
except MAX-kSAT, Hyper-Reorder's slowest $r$ is faster than the fastest $r$
of any other configuration.

@@

Previous work~\cite{ochoa:2015:crossovernetworks} noted that counter-intuitively Adjacent
NKq has more local optima than Random NKq for $r=1$ and $N=20$, even though the former is generally
easier to solve. However, many of these optima are ``weak'' in the sense that they are not also
2-bit local optima. Here we extend those conclusions for up to $N=59$ and up to $r=4$. Adjacent
NKq tend to have more 1-bit local optima than Random NKq, most of which disappear as $r$
increases. Furthermore, the growth rate of local optima between the two landscapes diverges
as $r$ increases.

MAX-kSAT and Ising Spin Glasses generally appear to contain far more local optima than
any of the other problems we enumerated. With just $N=49$, Ising Spin Glasses average
nearly 800 million local optima. For comparison, Adjacent NKq with $N=49$ averages fewer than 300 thousand,
and with $N=78$ Adjacent NKq still only has fewer than 500 million.
As with NKq, the polynomially solvable Ising Spin
Glass problem had more 1-bit local optima than the NP-Hard MAX-kSAT. Unlike with NKq, this
relationship generally does not change as $r$ increases. Even for $r=4$ the general
number and growth rate of local optima appears to stay the same for both problems.

@@

In our original experiment we tested problems up to size $N=100$, which was sufficient to
cause all three methods using $r=1$ to fail all problems except for Hyper-Reorder on Concatenated
Traps. However, using higher $r$ values, Hyper-Reorder is able to significantly extend its range.
On Random NKq, Hyper-Reorder was successful at finding 2-bit local optima for $N=73$. On Adjacent
NKq using all $r$ tested except 1 were successful on $N=100$. To push Algorithm~\ref{alg-enumerate}
to the extreme we tested Adjacent NKq with $N=200$ and $r=4$, with all 30 runs successful. On
average these runs required 5.6 minutes to complete. However, there was significant variance
in both the runtime and number of local optima. 18 of 30 runs completed in under 1 minute,
while the slowest run took 1.5 hours. 7 had less than one thousand 4-bit local optima, while 5
had over one hundred thousand.

>>>>

\subsection{All Optimal Orderings are Greedy}

@@

\subsection{$k$-Bound Separable Problems}
A $k$-bound separable problem is any problem that is composed of non-overlapping
subfunctions, each using using $k$ or fewer bits. Formally this means
$\forall_i |s_i| \leq k$ and $\forall_{i \neq j} s_i \cap s_j = \emptyset$.
While there is no restriction on how variables are ordered in the problem,
performing reordering will ensure that
all variables which appear in the same $s_i$ are consecutively placed during enumeration.
As such it is possible to create a numbering of $f_i$ such that if $i < j$ then all of $s_i$'s
variables appear before $s_j$'s in enumeration ordering.

Algorithm~\ref{alg-enumerate} finds the $f_i$ with the highest $i$ which contains a
fitness-improving move and enumerates its variables until it no longer
contains an improving move. In total, enumerating $f_i$ requires $2^k$
steps. Each $f_i$ is considered sequentially until none contains an improving move,
meaning the first local optimum is found in \BigO{2^kN/k} time.

Once the first local optimum is found, Algorithm~\ref{alg-enumerate} proceeds
by finding all local optimum in larger and larger hyperplanes. Initially $f_0$
is enumerated to find all local optima in the hyperplane with all $f_i, i>0$
fixed. This process ends when a carry operation modifies a bit of $f_1$. At
this point $f_1$ is enumerated, such that each time $f_1$ contains no improving moves
$f_0$ is enumerated again. This finds all local optima in the hyperplane with $f_i, i>1$
fixed. We represent the time required to find all local optima in the hyperplane
with $f_j, j>i$ fixed as $T(i)$. $T(0)=2^k$ as all values of $f_0$ must be enumerated.
$T(i) = |l_i|*T(i-1)+2^k$ where $|l_i|$ is the number of ways $f_i$ can be set such
that it contains no fitness-improving moves.
Each setting in $l_i$ exposes a new hyperplane
that can contain local optima, which causes the recursive call to $T(i-1)$.
As all $|s_i| \leq k$ and $k$ is a
constant, $\forall_i |l_i| \leq c \leq 2^k$ meaning that $c$ is a constant.
Assuming for simplicity that all subfunctions use exactly $k$ bits,
the time required to find all local optima is
$T(N/k-1)\leq \sum_0^{N/k-1}2^kc^i<2^kc^{N/k}$.
The number of local optima in the landscape is $c^{N/k}$ meaning that Algorithm~\ref{alg-enumerate}
is within a constant factor ($2^k$) of optimal.

@@

\subsubsection{Importance of Reordering}
The order in which variables are enumerated for $k$-bound separable problems
has a significant impact on Algorithm~\ref{alg-enumerate}'s complexity.
Instead of using the normal reordering procedure, consider an ordering
such that if $i < j$, all variables in $s_i$ appear before the second variable
in $s_j$ and after the first variable in $s_j$.

Each time the second variable in $s_i$ is flipped, all of $s_{i-1}$ must be enumerated.
This is because only carry operations can result in $index$ being set to the second
variable in $s_i$. Each time $f_i$ contains a fitness improving move, $index$ is set
to the first variable in $s_i$, which comes before all variables in $s_{i-1}$.
As a result the time to solve this ordering becomes $T(i) = 2^{k-1}*T(i-1)+2^k$.
Therefore, if $c < 2^{k-1}$ this ordering performs much worse than optimal.

@@

\begin{table}
	\centering
	\caption{Value of $m$ when fitting a model to $y = 2^c2^{mN}$ where $y$ is number of seconds required to
	         find all 1-bit local optima of a problem using $N$ bits. Optimal refers to the same model where
	         $y$ is the number of local optima. All $R^2$ values over 0.95 except Hyper on Concatenated Traps
	         which has 0.898.}
	\begin{tabular}{|r|c|c|c|c|}
	  \hline
	    & \textbf{Gray-Box} & \textbf{Hyper} & \textbf{Hyper-Reorder} & \textbf{Optimal} \\ \hline
    Concatenated Traps & 0.9809 & 0.5831 & 0.1620 & 0.2000 \\ \hline
    Adjacent NKq & 0.9922 & 0.5972 & 0.3419 & 0.3603 \\ \hline
    Random NKq & 0.9920 & 0.6229 & 0.3989 & 0.3531 \\ \hline
    Ising Spin Glass & 0.9821 & 0.6722 & 0.5976 & 0.6015 \\ \hline
    MAX-kSAT & 0.9949 & 0.8922 & 0.7737 & 0.5393 \\ \hline
  \end{tabular}
  \label{table-scaling}
\end{table}

With the exception of MAX-kSAT, Hyper-Reorder's time required to find all local optima
appears to be growing at the same rate as the total number of local optima. We suspect
some of this deficiency on MAX-kSAT is due to its significantly higher number of non-linear
interactions. Concatenated Traps, Adjacent NKq, and Ising Spin Glass all have exactly 4
non-linear relationships per variable. For small problem sizes
the number of non-linear relationships per variable in
Random NKq and MAX-kSAT both increase with problem
size. Random NKq's average ranges between 5 and 6 on problems we tested.
With $N=15$, MAX-kSAT has on average nearly
12 non-linear relationships per variable. By $N=30$ that grows to just over 17. Having each
variable depend on over half the genome significantly limits the size of hyperplanes which
can be eliminated. While the number of non-linear relationships per variable
is asymptotically constant, this growth on small problems is likely affecting runtime.

\subsection{Finding $r$-Bit Local Optima}
A major advantage of Gray-Box enumeration is that it can find $r$-bit local optima
with only a constant increase in runtime. Figure~\ref{fig-slope-method}
shows the estimated growth rate for each method as $r$ increases. On Concatenated Traps,
Adjacent NKq, Ising Spin Glass, and to a lesser extent Random NKq, none of our enumeration
methods see an increase in growth complexity with increasing $r$. In all cases using Hyper-Reorder
has a lower growth rate than the other two methods using the same $r$. On every problem
except MAX-kSAT, Hyper-Reorder's slowest $r$ is faster than the fastest $r$
of any other configuration.

@@

Previous work~\cite{ochoa:2015:crossovernetworks} noted that counter-intuitively Adjacent
NKq has more local optima than Random NKq for $r=1$ and $N=20$, even though the former is generally
easier to solve. However, many of these optima are ``weak'' in the sense that they are not also
2-bit local optima. Here we extend those observations for up to $N=59$ and up to $r=4$. Adjacent
NKq tends to have more 1-bit local optima than Random NKq, most of which disappear as $r$
increases. Furthermore, the growth rate of local optima between the two landscapes diverges
as $r$ increases.

MAX-kSAT and Ising Spin Glasses generally appear to contain far more local optima than
any of the other problems we enumerated. With just $N=49$, Ising Spin Glasses average
nearly 800 million local optima. For comparison, Adjacent NKq with $N=49$ averages fewer than 300 thousand,
and with $N=77$ Adjacent NKq still only has fewer than 500 million.
As with NKq, the polynomially solvable Ising Spin
Glass problem had more 1-bit local optima than the NP-Hard MAX-kSAT. Unlike with NKq, this
relationship generally does not change as $r$ increases. However, both do still
see large decreases in the number and growth rate of local optima. With $N=49$
the number of 4-bit local optima Ising Spin Glass is on average under 70 thousand,
a reduction of four orders of magnitude.

@@

In our original experiments we tested problems up to size $N=100$, which was sufficient to
cause all three methods using $r=1$ to fail all problems except for Hyper-Reorder on Concatenated
Traps. However, using higher $r$ values, Hyper-Reorder is able to significantly extend its range.
On Random NKq, Hyper-Reorder was successful at finding 2-bit local optima for $N=73$. On Adjacent
NKq using all $r$ tested except 1 were successful on $N=100$. To push Algorithm~\ref{alg-enumerate}
to the extreme we tested Adjacent NKq with $N=200$ and $r=4$, with all 30 runs successful. On
average these runs required 5.6 minutes to complete. However, there was significant variance
in both the runtime and number of local optima. Eighteen runs completed in under 1 minute,
while the slowest run took 1.5 hours. Three had fewer than 50 4-bit local optima, while five
had over 100,000.
